Working with daily climate model output data in R and the \pkg{futureheatwaves} package

### 2016-12-29 paper_drafts/RJournalRevisions/anderson-eason-barnes.Rmd
Work on R paper draft

\begin{figure}
\begin{center}
\includegraphics[width = 0.9\textwidth]{netcdf_structure_example}
\end{center}
\caption{Example of structure of a NetCDF climate model output file for a variable reported at a single depth, like surface air temperature. Data are stored in a three-dimensional array, with measurements at each time step and grid location. Surface temperature data are typically indexed in climate model output files by longitude, latitude, and time, in that order. For example, if the air surface temperature ("tas") is read into an R object called \code{tas}, you can access the value for the first day at the fourth longitude and third latitude with \code{tas[4, 3, 1]}. In addition to the output variable (temperature in this example), vectors with the ordered values of each dimension (longitude, latitude, and time) can also be read in from the netCDF file, as well as attribute data (e.g., units for variables, the calendar used for time).}
\label{fig:netcdfexample}
\end{figure}

@@

The netCDF format allows you to access metadata and variables describing the dimensions of the data without reading the full file into memory. The metadata describes the dimensions of the array and the units, etc., of each variable and is printed by the print method of the R object returned by the \code{nc\_open} function of \pkg{ncdf4}. For CMIP5 files, the dimension variables include the times and locations corresponding to each array in the data. Before reading in a variable from a netCDF file, a connection to the file must be opened, for example with the \code{nc\_open} function from \pkg{ncdf4}. Variables can then be read in using the \code{ncvar\_get} function from \pkg{ncdf4}, with the \code{varid} parameter set to "lat", "lon", or "time" (as a caveat, many climate models output to non-Gregorian calendars, in which case the time variable should be read in using a different function, as discussed later in this section). The climate output variable (e.g., near-surface air temperature) can similarly be read in using \code{ncvar\_get}. In this case, the \code{varid} parameter should be set using the appropriate CMIP5 variable name (e.g., "tas" for near-surface air temperature); these variable names can be found in the CMIP Requested Output tables \citep{taylor2010cmip5}. In practice, you can use the dimensional time and location data to identify the location of the variable data you need in the netCDF array and use indexing to read only that data into memory, without needing to read in the full file \citep{michna2013rnetcdf}. Once the user is done reading in data from the file, the connection can be closed (e.g., with the \code{nc\_close} function from \pkg{ncdf4}).

@@

The following code gives an example of using a CMIP5 file and some of the R packages and functions discussed. Results include a map of near-surface air temperatures from a single climate model on a specific date (July 1, 2075) for China (Figure [x]) and a time series of daily near-surface air temperature simulations for the climate model grid point closest to Beijing, China (Figure [x]). To try this example, you will need to download the file ... from ... and save it in the "tmp" subdirectory of your home directory.

```{r}

```

>>>>

\begin{widefigure}
\begin{center}
\includegraphics[width = \textwidth]{netcdf_figure}
\end{center}
\caption{Example of structure of a NetCDF climate model output file for a variable reported at a single depth, like surface air temperature. Data are stored in a three-dimensional array, with measurements at each time step and grid location. Surface temperature data are typically indexed in climate model output files by longitude, latitude, and time, in that order. For example, if the air surface temperature ("tas") is read into an R object called \code{tas}, you can access the value for the first day at the fourth longitude and third latitude with \code{tas[4, 3, 1]}. In addition to the output variable (temperature in this example), vectors with the ordered values of each dimension (longitude, latitude, and time) can also be read in from the netCDF file, as well as attribute data (e.g., units for variables, the calendar used for time).}
\label{fig:netcdfexample}
\end{widefigure}

@@

The netCDF format allows you to access metadata and variables describing the dimensions of the data without reading the full file into memory. The metadata describes the dimensions of the array and the units, etc., of each variable and is printed by the print method of the R object returned by the \code{nc\_open} function of \pkg{ncdf4}. For CMIP5 files, the dimension variables include the times and locations corresponding to each array in the data. Before reading in a variable from a netCDF file, a connection to the file must be opened, for example with the \code{nc\_open} function from \pkg{ncdf4}. Variables can then be read in using the \code{ncvar\_get} function from \pkg{ncdf4}, with the \code{varid} parameter set to "lat", "lon", or "time" (as a caveat, many climate models output to non-Gregorian calendars, in which case the time variable should be read in using a different function, as discussed later in this section). The climate output variable (e.g., near-surface air temperature) can similarly be read in using \code{ncvar\_get}. In this case, the \code{varid} parameter should be set using the appropriate CMIP5 variable name (e.g., "tas" for near-surface air temperature); these variable names can be found in the CMIP Requested Output tables \citep{taylor2010cmip5}. You can use the dimensional time and location data to identify the location of the variable data you need in the netCDF array and use indexing to read only that data into memory, without needing to read in the full file (for example, with the \code{nc.get.var.subset.by.axes} function in \pkg{ncdf4.helpers}). Once the user is done reading in data from the file, the connection can be closed (e.g., with the \code{nc\_close} function from \pkg{ncdf4}).

@@

The following code gives an example of using a CMIP5 file and some of the R packages and functions discussed. Results include a map of near-surface air temperatures from a single climate model on a specific day in July 2075 (Figure \ref{fig:worldmap}) and a time series of daily near-surface air temperature simulations for the climate model grid point closest to Beijing, China (Figure \ref{fig:timeseries}). To try this example, you will need to download the file "tas_day_GFDL-ESM2G_rcp85_r1i1p1_20710101-20751231.nc" from https://pcmdi.llnl.gov/search/cmip5/ and save it in the "tmp" subdirectory of your home directory.

```{r eval = FALSE, echo = FALSE}
# To try this example,you will need to download the file 
# “tas_day_GFDL-ESM2G_rcp85_r1i1p1_20710101-20751231.nc”
# from https://pcmdi.llnl.gov/search/cmip5/and save it in the “tmp” 
# subdirectory of your home directory

# Load required packages
# Packages for working with CMIP5 files
library(ncdf4)
library(ncdf4.helpers)
library(PCICt)
# Some general purpose packages
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggmap)
library(viridis)
library(weathermetrics)

# Open connection to CMIP5 netCDF file
climate_filepath <- paste0("~/tmp/tas_day_GFDL-ESM2G_rcp85",
                           "_r1i1p1_20710101-20751231.nc")
climate_output <- nc_open(climate_filepath)

# Load dimension variables (longitude, latitude, time) and climate output
# variable ("tas", near-surface air temperature)
lon <- ncvar_get(climate_output, varid = "lon")
lat <- ncvar_get(climate_output, varid = "lat")
tas_time <- nc.get.time.series(climate_output, v = "tas",
                               time.dim.name = "time")

# Subset to a single date (July 1, 2075) to create map
pdf("worldmapexample.pdf", height = 3, width = 6)
time_index <- which(format(tas_time, "%Y-%m-%d") == "2075-07-15")
tas <- nc.get.var.subset.by.axes(climate_output, "tas",
                                 axis.indices = list(T = time_index))
expand.grid(lon, lat) %>%
        rename(lon = Var1, lat = Var2) %>%
        mutate(lon = ifelse(lon > 180, -(360 - lon), lon),
               tas = as.vector(tas)) %>% 
        mutate(tas = convert_temperature(tas, "k", "c")) %>%
        ggplot() + 
        geom_point(aes(x = lon, y = lat, color = tas),
                   size = 0.8) + 
        borders("world", colour="black", fill=NA) + 
        scale_color_viridis(name = "Temperature (C)") + 
        theme_void() + 
        coord_quickmap() + 
        ggtitle("Modeled temperature on a day in July 2075",
                subtitle = "GFDL-ESM2G model, RCP8.5 experiment, r1i1p1 ensemble member") 
dev.off()

# Subset the time series for the grid point closest to Beijing, China, to
# create figure
pdf("timeseriesexample.pdf", height = 3, width = 6)
lon_index <- which.min(abs(lon - 116.4))
lat_index <- which.min(abs(lat - 39.9))
tas <- nc.get.var.subset.by.axes(climate_output, "tas",
                                 axis.indices = list(X = lon_index,
                                                     Y = lat_index))
data_frame(time = tas_time, 
           tas = as.vector(tas)) %>%
        mutate(time = as.Date(format(time, "%Y-%m-%d"))) %>%
        ggplot(aes(x = time, y = tas)) + 
        geom_line() + 
        xlab("Date") + ylab("Temperature (K)") + 
        ggtitle("Daily modeled near-surface air temprature, 2071-2075",
                subtitle = "At model grid point nearest Beijing, China") + 
        theme_classic()
dev.off()

# Close file connection to netCDF file
nc_close(climate_output)
```

\begin{figure}
\begin{center}
\includegraphics[width = \textwidth]{worldmapexample}
\end{center}
\caption{Example of mapping near-surface air temperature data worldwide for a single day of climate model output data. This map uses data from the Geophysical Fluid Dynamics Laboratory's Earth System Model 2G, r1i1p1 ensemble member, on a single day in the summer of 2075.}
\label{fig:worldmap}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width = 0.8\textwidth]{timeseriesexample}
\end{center}
\caption{Example of plotting a time series of temperature simulations between 2071 and 2075 from CMIP5 daily climate model output data for the model grid cell point closest to Beijing, China. This plot uses data from the Geophysical Fluid Dynamics Laboratory's Earth System Model 2G, r1i1p1 ensemble member.}
\label{fig:timeseries}
\end{figure}

### 2016-12-29 paper_drafts/RJournalRevisions/anderson-eason-barnes.tex
Work on R paper draft

The netCDF format allows you to access metadata and variables describing
the dimensions of the data without reading the full file into memory.
The metadata describes the dimensions of the array and the units, etc.,
of each variable and is printed by the print method of the R object
returned by the \code{nc\_open} function of \pkg{ncdf4}. For CMIP5
files, the dimension variables include the times and locations
corresponding to each array in the data. Before reading in a variable
from a netCDF file, a connection to the file must be opened, for example
with the \code{nc\_open} function from \pkg{ncdf4}. Variables can then
be read in using the \code{ncvar\_get} function from \pkg{ncdf4}, with
the \code{varid} parameter set to ``lat'', ``lon'', or ``time'' (as a
caveat, many climate models output to non-Gregorian calendars, in which
case the time variable should be read in using a different function, as
discussed later in this section). The climate output variable (e.g.,
near-surface air temperature) can similarly be read in using
\code{ncvar\_get}. In this case, the \code{varid} parameter should be
set using the appropriate CMIP5 variable name (e.g., ``tas'' for
near-surface air temperature); these variable names can be found in the
CMIP Requested Output tables \citep{taylor2010cmip5}. In practice, you
can use the dimensional time and location data to identify the location
of the variable data you need in the netCDF array and use indexing to
read only that data into memory, without needing to read in the full
file \citep{michna2013rnetcdf}. Once the user is done reading in data
from the file, the connection can be closed (e.g., with the
\code{nc\_close} function from \pkg{ncdf4}).

@@

The following code gives an example of using a CMIP5 file and some of
the R packages and functions discussed. Results include a map of
near-surface air temperatures from a single climate model on a specific
date (July 1, 2075) for China (Figure {[}x{]}) and a time series of
daily near-surface air temperature simulations for the climate model
grid point closest to Beijing, China (Figure {[}x{]}). To try this
example, you will need to download the file \ldots{} from \ldots{} and
save it in the ``tmp'' subdirectory of your home directory.

>>>>

The netCDF format allows you to access metadata and variables describing
the dimensions of the data without reading the full file into memory.
The metadata describes the dimensions of the array and the units, etc.,
of each variable and is printed by the print method of the R object
returned by the \code{nc\_open} function of \pkg{ncdf4}. For CMIP5
files, the dimension variables include the times and locations
corresponding to each array in the data. Before reading in a variable
from a netCDF file, a connection to the file must be opened, for example
with the \code{nc\_open} function from \pkg{ncdf4}. Variables can then
be read in using the \code{ncvar\_get} function from \pkg{ncdf4}, with
the \code{varid} parameter set to ``lat'', ``lon'', or ``time'' (as a
caveat, many climate models output to non-Gregorian calendars, in which
case the time variable should be read in using a different function, as
discussed later in this section). The climate output variable (e.g.,
near-surface air temperature) can similarly be read in using
\code{ncvar\_get}. In this case, the \code{varid} parameter should be
set using the appropriate CMIP5 variable name (e.g., ``tas'' for
near-surface air temperature); these variable names can be found in the
CMIP Requested Output tables \citep{taylor2010cmip5}. You can use the
dimensional time and location data to identify the location of the
variable data you need in the netCDF array and use indexing to read only
that data into memory, without needing to read in the full file (for
example, with the \code{nc.get.var.subset.by.axes} function in
\pkg{ncdf4.helpers}). Once the user is done reading in data from the
file, the connection can be closed (e.g., with the \code{nc\_close}
function from \pkg{ncdf4}).

@@

The following code gives an example of using a CMIP5 file and some of
the R packages and functions discussed. Results include a map of
near-surface air temperatures from a single climate model on a specific
day in July 2075 (Figure \ref{fig:worldmap}) and a time series of daily
near-surface air temperature simulations for the climate model grid
point closest to Beijing, China (Figure \ref{fig:timeseries}). To try
this example, you will need to download the file
``tas\_day\_GFDL-ESM2G\_rcp85\_r1i1p1\_20710101-20751231.nc'' from
\url{https://pcmdi.llnl.gov/search/cmip5/} and save it in the ``tmp''
subdirectory of your home directory.

\begin{figure}
\begin{center}
\includegraphics[width = \textwidth]{worldmapexample}
\end{center}
\caption{Example of mapping near-surface air temperature data worldwide for a single day of climate model output data. This map uses data from the Geophysical Fluid Dynamics Laboratory's Earth System Model 2G, r1i1p1 ensemble member, on a single day in the summer of 2075.}
\label{fig:worldmap}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width = 0.8\textwidth]{timeseriesexample}
\end{center}
\caption{Example of plotting a time series of temperature simulations between 2071 and 2075 from CMIP5 daily climate model output data for the model grid cell point closest to Beijing, China. This plot uses data from the Geophysical Fluid Dynamics Laboratory's Earth System Model 2G, r1i1p1 ensemble member.}
\label{fig:timeseries}
\end{figure}
